\chapter{Bibliography}

Aho, A.V., R. Sethi and JD. Ullman, \textit{Compilers: ...} (Addison-Wesley. Readlng. 1988).

Bamhart. Joe "FORTH and the Fast Fourier Transform" \textit{Dr. Dobb's Journal}. September, 1984. p. 34.

Basie. F.J. \textit{J. FORTH Appl. and Res.} 1.2 (1982) 76-78.

Berestetk\widetilde{ii}. V.B., EM. Lifshitz and L.P. Pitaevsk\widetilde{ii}. \textit{Relativistic Quantum Theory, Part 1} (Pergamon Press. Oxford. 1971) p. 68ff.

Berrian. D.W. \textit{Proc. 1989 Rochester FORTH Conf.} (Inst. for Applied FORTH Res., Inc., Rochester, NY 1989) p. 1-5.

Bjorken, J.D. and S.D. Drell. \textit{Relativistic Quantum Mechanics (McGraw-Hill, Inc., New York, 1864)

Bratley, P., B.L. Fox and L.E. Schrage, \textit{A Guide to Simulation} (Springer-Verlag, Berlin, 1983).

Brodie. L. \textit{Starting FORTH}, 2nd ed. (PremioeHall. NJ. 1986)
Brodie. L. \textit{Thinking Forth} (Prentice-Hall. NJ. 1984).

Callahan, J.S. \textit{Proc. 1988 Rochester FORTH Conference} (Inst for Applied FORTH Research, Inc., 1988), p. 39.

Clark, D.D. "Simple Calculations with Complex Numbers", \textit{Dr. Dobb's Journal}, October 1984, p. 30.

Crawford, John H. and Patrick P. Gelsinger, \textit{Programming the 80386} (Sybex, San Francisco, 1987).

Duncan, Ray "FORTH support for Intel/Lotus expanded memory", \textit{Dr. Dobb's Joumal}, August 1986.

Duncan, Ray and Martin Tracy, \textit{Dr. Dobb's Jounal}, September 1984, p. 110.

Gedeon, D. "Complex Math in Pascal", \textit{Byte Magazine}, July 1987, p. 121.

Ham, M. “Structured Programming", \textit{Dr. Dobb's Joumal}, July, 1986.
Ham, M. “Structured Programming", \textit{Dr. Dobb's Joumal}, October 1986.

Hammersley, J.M. and D.C. Hanscomb, \textit{Monte Carlo Methods} (Methuen, London, 1964).

\textit{Handbook of Mathematical Functions} ed. Milton Abramowitz and Irene Stegun (Dover Publications, Inc., New York, 1965).

Harvard Softworks, P.O. Box 69, Springboro, Ohio 45066 Tel: (513) 748-0390.

Ivanco, T.A. and G. Hunter, \textit{J. Forth Appl. and Res.} \textbf{6} (1990) 15.

Jones. Do-While \textit{Dr. Dobb's Journal}, March 1987.

Jump, D.N. \textit{Programmer's guide to MS-DOS}, rev. ed. (Brady Books. New York, 1987).

Kelly, M. and N. Spies, \textit{FORTH: a Text and Reference} (Prentice-Hall, NJ, 1986)

Knuth, D.E. \textit{The Art of Computer Programming, v. 2: Seminumerical Algorithms}, 2nd ed. (Addison-Wesley Publishing Company, Reading, MA 1981)

Kohavi, Zvi \textit{Switching and Finite Automata Theory}, 2nd ed. (McGraw-Hill Publishing Co., New York, 1978).

Kruse, R.L. \textit{Data Structures and Program Design}, 2nd ed. (Prentice-Hall, Inc., Englewood Cliffs, NJ 1987).

Lafore, R. \textit{Assembly Language Primer for the IBM PC & XT} (Plume/Waite -New American Library, New Your, 1984).

Lefor, John A. and Karen Lund, "Reaching into expanded memory", \textit{PC Tech Journal}, May 1987.

Loeliger, R.G. \textit{Threaded Interpretive Languages} (Byte Publications, Inc., Peterborough, NH, 1981).

Mathews, J. and R.L. Walker, \textit{Mathematical Methods of Physics, 2nd ed. (W.A. Benjamin, Inc., New Jersey, 1970).

Mignotte, M. \textit{Mathematics for Computer Algebra} (Springer-Verlag, Berlin, 1992).

Morgan, C. and M. Waite, \textit{8086/8088 16-bit microprocessor primer} (Byte/McGraw-Hill, Peterborough, 1982).

Noble, J.V. "Avoid Decisions", \textit{Computers in Physics} \textbf{5}, 4 (1991) 396.
Noble, J.V. "Scientific Computation in FORTH", \textit{Computers in Physics} \textbf{3} (1989) 31.
Noble, J.V. \textit{J. FORTH Ap. and Res.} \textbf{6} (1990) 47.
Noble, J.V. \textit{J. FORTH Ap. and Res.} \textbf{6} (1990) 131.

Palmer, J.F. and S.P. Morse, \textit{The 8087 Primer} (John Wiley and Sons, Inc., New York, 1984).

Pan. V. \textit{SIAM Review} \textbf{26} (1984) 393.

Pavelle, R., M. Rothstein and J. Fitch, "Computer Algebra", \textit{Scientific American} \textbf{245}, \#6 (Dec. 1981) 136.

Pountain, Dick "Object-oriented FORTH", Byte Magazine, 8/86; \textit{Object-oriented FORTH} (Academic Press,
Inc., Orlando, 1987).

Press, W.H., B.P. Flamery, S.A. Teukolsky and WT. Vetterling, \textit{Numerical Recipes} (Cambridge University Press, Cunbrﬂge. 1986)

Ralston, A. \textit{A First Course in Nunerical Analysis} (McGraw-Hill Book Company, New York, 1965).

Rawson, E. \textit{J. FORTH Appl. and Res.} \textbf{3,4} (1986) 45-64.

Scanlon, L.J. \textit{IBM PC & XT assembly language: a guide for programmers} (Brady/Prentice-Hall, Bowie, Md., 1983).

Sedgewick, R. \textit{Algorithms} (Addison-Wesley Publishing Company, Reading, MA, 1983).

Shaw, G. "Forth Shite Gears, I", \textit{Computer Language} (May 1988) p. 67
Shaw, G. "Forth Shite Genre, ll", \textit{Computer Language} (June 1988) p. 61
Shaw, G. \textit{Proc. MAeilomar FORML Conference} (JFAR \textbf{5} (1988) 347.)

Strassen, V. textit{Numer. Mathe.} \textbf{13} (1969) 184.

Williams, Al "DOS + 386 = 4 Gigabytes", \textit{Dr. Dobb's Journal}, July 1990, p. 62.